k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-14 paper.Rmd
minor

The *k*-plus criterion is an extension of the $k$-means criterion for anticlustering that can be implemented by adding fictitious variables to the data matrix $\mathbf{X}_{N \times M}$ [@papenberg2024]. *K*-plus anticlustering is used to equate not only means between clusters, but higher order moments such as the variance or skewness as well. For example, to equate means and variances, we define a new data matrix $\mathbf{Y}_{N \times M} = \{y_{ij}\}_{N \times M}$ with $y_{ij} = (x_{ij} - \overline{x}_j)^2$, where $\overline{x}_j$ is the overall mean of the $j^{th}$ attribute, and augment the original matrix with it. Then, the *k*-plus criterion is computed as

@@

The diversity is oftentimes obtained by first converting the data matrix $\mathbf{X}_{N \times M}$ into a matrix of pairwise Euclidean distances [@brusco2019; @papenberg2020; @papenberg2024; @gallego2013]. If $\mathbf{D}_{N \times N}$ is obtained by using the squared pairwise Euclidean distances instead, we note an important equivalence between the k-means criterion $f_1$ and the diversity $f_3$. Traditionally, the k-means criterion is denoted via a sum of squared distances between data points and cluster centroids. However, it can also be expressed using the pairwise squared Euclidean distances between data points (Brusco, 2006; Späth, 1980). If we let $\mathbf{D'}_{N \times N}$ = \{$d'_{ij}$\} representing the squared Euclidean distance, i.e., $d'_{ij} = \sum\limits_{m = 1}^{M}(x_{im} - x_{jm})^2$, the k-means criterion is given as

@@

That is, "the sum of squared distances of a collection of points from the centroid associated with those points is equal to the sum of the pairwise squared distances between those points divided by the size (number of objects) of the collection" (Brusco, 2006, p. 350). An important special case of anticlustering applications is that all groups are equal-sized, i.e., $n_k = \frac{N}{K}$ ($k = 1, \dots K$). In this context, $f_4$ can be simplified to $f^*$ via

$$
f_4^*(C, \mathbf{D'}_{N \times N}) = 
  \frac{N}{K} \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij} \\
  \propto \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij} = f_3(C, \mathbf{D'}_{N \times N})
$$

In the context of maximization, the normalizing factor $\frac{N}{K}$ can be ignored because it does not depend on the partitioning $C$. Thus, the criterion $f_4^*$ reduces to the sum of the (squared) distances between elements in the same cluster, i.e., the standard diversity $f_3$. Therefore, for equal-sized groups, the diversity anticlustering criterion $f_3$ is equivalent to the k-means criterion $f_1$---if the diversity is computed via the squared Euclidean distance. We will use this insight later when we discuss applying algorithms for maximum diversity to the $k$-means and $k$-plus objectives. For arbitrary measures of dissimilarity (and not just the squared Euclidean distance), we denote $f_4$ as the *average diversity criterion*.

@@

@brusco2019 outlined the importance of simultaneously considering multiple criteria when tackling anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity---such as the diversity---and pairwise within-group dissimilarity---i.e., the dispersion---at the same time. To this end, @brusco2019 presented a bicriterion algorithm that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$. (**say some general stuff about the BILS before explaining it in detail later...**)

We endorse Brusco et al.'s bicriterion approach for anticlustering, and advance it in several ways in the current paper. Specifically, we offer several contributions: First, we allow that not only the standard diversity $f_3$ can be optimized alongside the dispersion, but instead we allow that the k-means, k-plus or the average diversity can be used as measures of between-group similarity (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion and $f_1$ -- $f_4$ can be optimized using different dissimilarity matrices. This is useful if both criteria should be based on different information, or to induce custom cannot-link constraints, as we will show. As out most important contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Despite its theoretical computational hardness, our optimal method scales to rather large data sets. It generates multiple partitions that we use as initializations of the BILS's local search, thus guaranteeing that the BILS also returns a partition with optimal dispersion. In a simulation study, we compared several adaptations of the BILS.

>>>>

The *k*-plus criterion is an extension of the $k$-means criterion for anticlustering that can be implemented by adding fictitious variables to the data matrix $\mathbf{X}_{N \times M}$ [@papenberg2024]. *K*-plus anticlustering is used to equate not only means between clusters, but higher order moments such as the variance or skewness as well. For example, to equate means and variances, we define a new data matrix $\mathbf{Y}_{N \times M} = \{y_{ij}\}_{N \times M}$ with $y_{ij} = (x_{ij} - \overline{x}_j)^2$, where $\overline{x}_j$ is the overall mean of the $j^{th}$ attribute. Then, the *k*-plus criterion is computed as

@@

The diversity is oftentimes obtained by first converting the data matrix $\mathbf{X}_{N \times M}$ into a matrix of pairwise Euclidean distances [@brusco2019; @papenberg2020; @papenberg2024; @gallego2013]. If $\mathbf{D}_{N \times N}$ is obtained by using the squared pairwise Euclidean distances instead, we note an important equivalence between the k-means criterion $f_1$ and the diversity $f_3$. Traditionally, the k-means criterion is denoted via a sum of squared distances between data points and cluster centroids. However, it can also be expressed using the pairwise squared Euclidean distances between data points (Brusco, 2006; Späth, 1980). If we let $\mathbf{D'}_{N \times N}$ = \{$d'_{ij}$\} represent the squared Euclidean distance, i.e., $d'_{ij} = \sum\limits_{m = 1}^{M}(x_{im} - x_{jm})^2$, the k-means criterion can be expressed as as

@@

That is, "the sum of squared distances of a collection of points from the centroid associated with those points is equal to the sum of the pairwise squared distances between those points divided by the size (number of objects) of the collection" (Brusco, 2006, p. 350). An important special case of anticlustering applications is that all groups are equal-sized, i.e., $n_k = \frac{N}{K}$ ($k = 1, \dots K$). In this context, $f_4$ can be simplified to $f_4^*$ via

$$
f_4^*(C, \mathbf{D'}_{N \times N}) = 
  \frac{N}{K} \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij}
$$

In the context of maximization, the normalizing factor $\frac{N}{K}$ can be ignored because it does not depend on the partitioning $C$. Hence,

$$
f_4^*(C, \mathbf{D'}_{N \times N})
  \propto \sum\limits_{k = 1}^{K} \sum_{(i<j)\in C_k} d_{ij} = f_3(C, \mathbf{D'}_{N \times N})
$$

For anticlustering, the criterion $f_4^*$ (i.e., the k-means criterion for equal-sized groups) reduces to the sum of the (squared) distances between elements in the same cluster, i.e., the standard diversity $f_3$. Therefore, for equal-sized groups, the diversity anticlustering criterion $f_3$ is equivalent to the k-means criterion $f_1$---if the diversity is computed via the squared Euclidean distance. We will use this insight later when we discuss applying algorithms for maximum diversity to the $k$-means and $k$-plus objectives. For arbitrary measures of dissimilarity (and not just the squared Euclidean distance), we denote $f_4$ as the *average diversity criterion*.

@@

@brusco2019 outlined the importance of simultaneously considering multiple criteria when addressing anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity---such as the diversity---and pairwise within-group dissimilarity---i.e., the dispersion---at the same time. To this end, @brusco2019 presented a bicriterion algorithm that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$. While in the next section, we will discuss their approach in more detail. We endorse Brusco et al.'s bicriterion approach for anticlustering, and advance it in several ways in the current paper. Specifically, we offer several contributions: First, we allow that not only the standard diversity $f_3$ can be optimized alongside the dispersion, but instead we allow that the k-means, k-plus or the average diversity can be used as measures of between-group similarity (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion and $f_1$ -- $f_4$ can be optimized using different dissimilarity matrices. This is useful if both criteria should be based on different information, or to induce custom cannot-link constraints, as we will show. As out most important contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Despite its theoretical computational hardness, our optimal method scales to rather large data sets. It generates multiple partitions that we use as initializations of the BILS's local search, thus guaranteeing that the BILS also returns a partition with optimal dispersion. In a simulation study, we compared several adaptations of the BILS.
